---
title: "Regression Models Course Project"
author: "John B Cheadle"
date: "October 17, 2017"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(ggplot2)
require(car)
```

# Executive Summary

This report explores the relationship between between a set of variables and miles per gallon (MPG) using the mtcars data set. Specifically, the repot seeks to answer two questions:

1. Is an automatic or manual transmission better for MPG?

2. What is the quantifiable difference between automatic and manual transmissions?

The initial linear model fitted with MPG as the otucome and automatic/manual as the predictor, **lm(mpg ~ factor(am), data=mtcars)**, shows an increase of 7.425 mpg switching from automatic to manual transmission.  However, there is a hidden variable, weight(wt), which is highly correlated with the am variable.  When we add wt to our linear model, **lm(mpg ~ factor(am) + wt, data=mtcars)**, we show a *decrease* of .024 mpg switching from automatic to manaul transmission, an example of Simpson's paradox. Therefore, we conclude that when weight is included as a regressor in our model, automatic/manual transmission is not a significant predictor for mpg in our mtcars data set.

# Exploring the mtcars data set
To start, we look at some aspects of the mtcars data set.

```{r mtcars_help, echo=FALSE}
?mtcars
data(mtcars)
head(mtcars)
sapply(mtcars, class)
```
Here we get a sense of the mtcars data set.  We learn that 'am' is one of the main variables we want to regress on (0 = automatic, 1 = manual).

```{r mtcars_plot, echo=FALSE}
boxplot(mpg~am,data=mtcars, 
        main="Car Milage Data",
        xlab="Transmission (0=auto, 1=manual)",
        ylab="Miles Per Gallon (MPG)")
```

It appears from the relative means of MPG for automatic and manaul transmission that these two variables are correlated.  In the following sections, we fit a linear model using lm() with the outcome 'mpg' and the primary regressor 'am' from the mtcars data set.

# Model Fitting
The initial model fitted used the variable am as the regressor to the mpg outcome.

```{r modelfit1, echo=TRUE}
fit1 <- lm(mpg ~ factor(am), data=mtcars)
summary(fit1)$coef
```
Here the intercept is the mean mpg for am value of 0 (automatic), whereas factor(am)1 is the increase in mpg switching to manual, which is 7.245 (a significant increase).  **So we are gaining 7.245 mpg from switching from automatic to manual, all other variables held constant.**

# Residual Plot & Diagnostics
To probe that the variation we see in our fit is not systemic, we investigate residuals.  Typically, to probe problems in the model, we plot residuals vs. fitted values.

Clearly, there is something systemic **(Figure A1, Appendix)** that is causing variation, hinting at a hidden variable.

We know intuitively that other factors influence a car's mpg, such as weight, # of cylinders, etc.  We explore the effect of weight on mpg in an exploratory plot, denoting whether the car is automatic or manual:

```{r mtcars_ggplot, echo=TRUE}
g = ggplot(data=mtcars, aes(wt,mpg))
g = g + geom_point(size=4, alpha=1/2,
                   aes(color=factor(am)))
g = g + ggtitle('Investigating mtcars Variables')
print(g)
```

Here things become more clear; it appears that the heavier cars are almost all automatic (0, pink), whereas the lighter cars are almost all manual transmission (1, blue).

# Model Fitting, cont.
In light of this new information, we fit more models and assess their fit using analysis of variance (ANOVA).

```{r modelfit2, echo=TRUE}
fit2 <- lm(mpg ~ factor(am) + wt, data=mtcars)
fit3 <- lm(mpg ~ factor(am) + wt + cyl, data=mtcars)
fit4 <- lm(mpg ~ ., data=mtcars)

anova(fit1, fit2, fit3, fit4)
```
From the p-values it appears that other variables are necessary to include (fit2), but only up to a point; if we include all of them, our p-value drops.  Here we decide to go with fit2, the model that includes the variable wt.

To investigate coefficients of this new model, we re-run the summary:

```{r modelfit_wt, echo=TRUE}
summary(fit2)$coef
```
We notice that our factor(am)1 reversed sign and is not a significant predictor of mpg when weight is included as a regressor - This is an example of Simpson's paradox.  We reinterpret our coefficients as such: when we include am and wt as regressors, a switch from automatic to manual transmission *decreases* mpg by 0.24, whereas a 1000 lb increase in car weight decreases our mpg by 5.35.


## Conclusions
Based on p-values from our ANOVA, we chose the **Model 2: mpg ~ factor(am) + wt** as the most appropriate fit.  In this model, we see a 0.24 *decrease* in mpg switching from automatic to manual.  We also see a 5.35 *decrease* in mpg per 1000lb of car weight.

To probe this model, we again plot the residuals vs the fitted (yhat) values.  As seen in Appendix figure A2, the scatterplot simply looks like a cloud of data points, suggesting an unbiased variation.

We also quantify the uncertainty of our chosen model.

```{r uncertainty, echo=TRUE}
sumCoef <- summary(fit2)$coefficients
sumCoef[2,1] + c(-1,1)*qt(.975,df=fit2$df) * sumCoef[2,2]
sumCoef[3,1] + c(-1,1)*qt(.975,df=fit2$df) * sumCoef[3,2]
```

The results: for our 1st coefficient 'am', a switch from automatic to manual results in a [-3.18 to 3.14] increase in mpg with 95% certainty.  This passes through 0 and is quite even on both sides, therefore insignificant.  For our 2nd coefficient 'wt', an increase in 1000lb of car weight is going to result in a [-6.96 to -3.74] increase in mpg, with 95% certainty.  This suggests am is insignificant, where increase in weight significantly reduces mpg.

In conclusion, our model suggests that manual/automatic transmission has *no effect* on mpg; rather, **weight is the primary predictor of mpg outcome.**

## Appendix
```{r Figure_A1, echo=TRUE}
yhat <- predict(fit1)
e <- resid(fit1)
plot(yhat, e, main = "Figure A1, Residuals vs. Fitted")
```

```{r Figure_A2, echo=TRUE}
yhat2 <- predict(fit2)
e2 <- resid(fit2)
plot(yhat2, e2, main = "Figure A2, Residuals vs. Fitted")
```

