---
title: "Regression Models Course Project"
author: "John B Cheadle"
date: "October 17, 2017"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Executive Summary
Q: 	8. Did the report include an executive summary?


This report explores the relationship between between a set of variables and miles per gallon (MPG) using the mtcars data set.  After some exploratory data analysis, I will use linear models to answer two primary questions:
1. Is an automatic or manual transmission better for MPG?

2. What is the quantifiable difference between automatic and manual transmissions?

Brief Results

## Exploring the mtcars data set
- Did the student do some exploratory data analyses?

```{r mtcars_help, echo=FALSE}
?mtcars
data(mtcars)
head(mtcars)
sapply(mtcars, class)
```
Here we get a sense of the mtcars data set.  We learn that 'am' is one of the main variables we want to regress on (0 = automatic, 1 = manual).

```{r mtcars_plot, echo=TRUE}
boxplot(mpg~am,data=mtcars, 
        main="Car Milage Data",
        xlab="Transmission (0=auto, 1=manual)",
        ylab="Miles Per Gallon (MPG)")
```
It appears from the relative means of MPG for automatic and manaul transmission that these two variables are correlated.  In the following sections, we fit a linear model using lm() with the outcome 'mpg' and the primary regressor 'am' from the mtcars data set.

## Model Fitting
- Did the student fit multiple models and detail their strategy for model selection?

## Residual Plot & Diagnostics
- Did the student do a residual plot and some diagnostics?

## Interpretation of Coefficients
- Did the student interpret the coefficients correctly?

## Conclusions
- Did the student answer the questions of interest or detail why the question(s) is (are) not answerable?

- Did the student quantify the uncertainty in their conclusions and/or perform an inference correctly?

## Appendix